Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-03-28 paper/paper.tex
Added Nick to author list.
Trying to reorder for better exposition.

\authorinfo{J. Ian Johnson}
           {Northeastern University}
           {ianj@ccs.neu.edu}
\authorinfo{Matthew Might}
           {University of Utah}
           {might@cs.utah.edu}
\authorinfo{David Van Horn}
           {Northeastern University}
           {dvanhorn@ccs.neu.edu}

\maketitle

@@

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{alloc}$, $\mathit{tick}$,
  $\mathit{push}$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
&&\mcntr' =\tick(\mcntr) \\
&&\maddr  =\alloc(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\push_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
$\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction that returns an abstract top element $\mathit{Integer}$ as
the result of arithmetic operations.  This family of interpreters is
also non-deterministic in choices of addresses, but it is free to
choose addresses that are already in use.  Consequently, the machines
may be non-deterministic when multiple values reside in a store
location.

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\push_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

The possible traces of execution tell us the most about a program, so
we take $\traces(\mexp)$ to be the (regular) set of paths through the
computed graph.

@@

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and it is deterministic even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

For clarity, we will exhibit non-widened semantics unless a technique
specifically requires it.

\subsection{Store-allocated results}
\label{sec:baselineeval}

The final approximation we make to get to our point of departure is
store-allocating results of application sub-expressions. 
The AAM approach stops at the previous optimization. 
However, the {\tt fn} continuation stores a value, and this makes the space of
continuations quadratic rather than linear in the size of the program---even for a monovariant analysis like 0-CFA.
Having the space of continuations grow linearly with the size of the program 
will drop the overall complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate addresses to store the values of each
subexpression of an application. The continuations have to store these
addresses to remember where to get argument values from, though
sometimes your allocation strategy may allow you to elide that
representation.

The new evaluation rules follow (where $\mcntr' = \tick(\mlab,\mcntr)$).
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

@@

\begin{align*}
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar{\mexpi{1},\menv,\maddr_f,\maddr_a,\maddr_\mkont}} \\
\text{ where }
  \maddr_\mkont,\msto' &= \push^\mcntr_\mlab(\msto,\mkont) \\
  \maddr_f,\maddr_a &= \allocsub^\mcntr_\mlab(\menv,\msto,\mkont)
\end{align*}
This rule describes how we allocate addresses for subexpressions of an
application up front. This adds another parameter to the machine, but
it is very much like $\alloc$. We use $f$ and $a$ subscripts to denote
function and argument positions, respectively. Whereas ANF names all
intermediate values in the syntax, we allocate addresses for
intermediate values. This representational leap is not a big one, but
is notably important for analyzing direct style terms.

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_a,\maddr_\mkont}} \\
\text{ where }
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\\
\co[^\mcntr]{\kfn{\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr_a,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f) \\
\text{ where }
 \msto' &= \ext{\msto}{\maddr_a}{\{\mval\}}
\end{align*}
We associate the value in hand with the expression that produced
it. Also, we fan out to all functions that are associated with the
original function position, because applied functions are in
\emph{strict position} (more about that later).

\begin{align*}
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\maddr_a, \msto,\mkont} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv',\msto',\mkont} \\
\text{ where }
\maddr &= \alloc(\mstate) \\
\menv' &= \menv[\mvar\mapsto\maddr] \\
\msto' &= \msto\sqcup[\maddr \mapsto \msto(\maddr_a)]
\end{align*}
Arguments are not passed as values, but as addresses to their values.

\begin{align*}
\ap[^\mcntr_\mlab]{\mop,\maddr_a,\msto,\mkont} & \machstep
\co[^{\mcntr'}]{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\maddr_a)\}
\end{align*}
Primitive application still nondeterministically produces all possible values.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) transition elimination
and (2) pragmatic improvement.
%
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
  \setcounter{enumi}{2}
 \item store deltas;
 \item timestamped stores; and
 \item preallocated data structures.
\end{enumerate}

All techniques are precision preserving under reasonable
assumptions. We will discuss these assumptions as the caveats arise.

@@

\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\mval) &= \{\mval\}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_f,\maddr_a,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_a,\maddr_\mkont}} \\
\text{ where }
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^\mcntr]{\mexpi,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\maddr_\mkont),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take. The middle rule uses $\force$
only to combine with values in the store - it does not introduce
needless non-determinism.

>>>>

\authorinfo{J. Ian Johnson}
           {Northeastern University}
           {ianj@ccs.neu.edu}
\authorinfo{Nicholas Labich}
           {Northeastern University}
           {labichn@ccs.neu.edu}
\authorinfo{Matthew Might}
           {University of Utah}
           {might@cs.utah.edu}
\authorinfo{David Van Horn}
           {Northeastern University}
           {dvanhorn@ccs.neu.edu}
\maketitle

@@

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

@@

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\maddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto \\
\allockont^\mcntr_\mlab(\msto,\mkont) \mbox{ where } \maddr \notin \msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
$\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction where arithmetic operations return an abstract top element
$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
$\sNum$.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

The possible traces of execution tell us the most about a program, so
we take $\traces(\mexp)$ to be the (regular) set of paths through the
computed graph. We elide the construction of the set of edges in this paper.

@@

To describe this process, AAM defined a transformation of the reduction relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The transformed relation, $\widehat{\machstep}$, is
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \mkont, \msto} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and it is deterministic even if set iteration is not.

\begin{align*}
(S, F, \msto) &\mathrel{\widehat{\machstep}} (S \cup S', F', \msto') \\
\mbox{where } S' &= \{(c',\msto^c) \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in F\} \\
              F' &= \{c \mid (c,\_) \in S' \setminus S\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid (\_,\msto^c) \in S'\} \\
\inject(e) &= (\setof{\ttuple{c_0}{\bot}},\setof{c_0},\bot) \\
\text{ where } c_0 &= \ev{e,\bot,\kmt}
\end{align*}

% Notice that now $S$ has several copies of the abstract store in it. As
% it is, this semantics is much less efficient (but still more precise)
% than the previously proposed semantics because membership checks have
% to compare entire stores. We return to this in section
% \ref{sec:timestamping}.

For clarity, we will exhibit non-widened semantics unless a technique
specifically requires it. The final step to the baseline takes the
complexity from quartic to cubic in the monovariant case.

\subsection{Store-allocate all values}
\label{sec:baselineeval}

The final approximation we make to get to our point of departure is to
store-allocate all values that appear, so that any non-machine state
that contains a value instead contains an address to a value.  The AAM
approach stops at the previous optimization.  However, the {\tt fn}
continuation stores a value, and this makes the space of continuations
quadratic rather than linear in the size of the program, for a
monovariant analysis like 0-CFA.  Having the space of continuations
grow linearly with the size of the program will drop the overall
complexity to cubic (as expected).

To achieve this linearity for continuations, we allocate an address
for the value position when we create the continuation.  This address
and the tail address are both determined by the label of the
application point, so the space becomes linear and the overall
complexity drops to cubic.  This is a critical abstraction in
languages with $n$-ary functions, since otherwise the continuation
space grows super-exponentially. We extend the semantics to
additionally allocate an address for the function value when creating
the $\kfn{}$ continuation. The continuation has to store this address
to remember where to get values from, though sometimes your allocation
strategy may allow you to elide that representation.

The new evaluation rules follow (where $\mcntr' = \tick(\mstate)$).
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.

@@

\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr_f,\maddr_\mkont}} \\
\text{ where }
  \maddr_f &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr_f}{\{\mval\}}
\\
\co[^\mcntr]{\kfn{\maddr_f,\maddr_\mkont},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\mval,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\maddr_\mkont), \mvalx{u} \in \msto(\maddr_f)
\end{align*}
We associate the value in hand with the expression that produced
it. Also, we fan out to all functions that are associated with the
original function position, because applied functions are in
\emph{strict position} (more about that later).

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, preallocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item lazy non-determinism and
 \item abstract compilation;
\end{enumerate}

All pragmatic improvements are precision preserving (form complete
abstractions), but the semantic changes are not in some cases, for reasons we
will describe. We did not observe the precision differences in our evaluation.

The move to the imperative will be made last in order to show the
effectiveness of these techniques in the purely functional realm.

\subsection{Timestamping monotonicity}

The frontier-based semantics of the baseline has a severe bottleneck:
checking if a state has been explored or not. Given two states,
checking equality is expensive because the stores within each are
large, and every entry must be checked against every other. Hashes can
sometimes rule out inequality relatively quickly, but the incidence of
collisions and actual equality is costly.

\begin{enumerate}
\item{In our semantics, the store does not change until the entire
    frontier has been explored. This avoids cross-branch pollution
    which would not otherwise happen, e.g., when one branch writes to
    address $\maddr$ and another branch reads from address
    $\maddr$.}
\item{The common implementation strategy for timestamps destructively
    updates each state's timestamp. This loses \emph{temporal}
    information about the contexts a state is visited in, and in what
    order. The semantics is different - we can't precisely answer
    questions about the traces of the reduction relation we defined.
    Our semantics has a drop-in replacement of timestamps for stores
    in the seen set ($S$), so we do not experience precision loss.}
\end{enumerate}

@@

\begin{align*}
\spchoice
{\force &: \Store \times \Value \to \Set(\Value) \\
 \force(\msto,\saddr{\maddr}) &= \msto(\maddr) \\
 \force(\mval) &= \{\mval\}}
{\force &: \Value \to \Set(\Value) \\
 \force(\superposition{\mval{s}}) &= \mval{s} \\
 \force(\mval) &= \{\mval\}}
\\
%
\ev{\svar{\mvar},\menv,\mkont,\msto} &\lmachstep\;
\spchoice
{\co{\mkont,\saddr{\menv(\mvar)},\msto}}
{\co{\mkont,\superposition{\msto(\menv(\mvar))},\msto}} \\
%
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^\mcntr]{\mexp,\menv,\msto',\kfn[^\mcntr_\mlab]{\maddr_f,\maddr_\mkont}} \\
\text{ where }
\maddr_f &= \alloc(\mstate) \\
\msto' &=
\spchoice
{\msto \sqcup[\maddr \mapsto \force(\msto,\mval)]}
{\msto \sqcup[\maddr \mapsto \force(\mval)]} \\
%
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\mval,\msto}
&\lmachstep\;
\ev[^\mcntr]{\mexpi,\menv,\msto,\mkont} \\
\text{ if } \mkont &\in \msto(\maddr_\mkont),
            \strue \in \spchoice{\force(\msto,\mval)}{\force(\mval)}
\end{align*}
Since {\tt if} guards are in strict position, we must force the value
to determine which branch to to take. The middle rule uses $\force$
only to combine with values in the store - it does not introduce
needless non-determinism.
